---
title: "suicide is a scary word"
output: html_document
---
##Introduction :
   Suicide or ending one's own life is a tragic event with strong emotional repercussions for its survivors and for families of its victims.Men are especially at risk, with a suicide rate approximately four times higher than that of women.
   So lets analyse the major causes for suicide attacks, which age group does that more in **INDIA**.
  

```{r echo=FALSE}

library(ggplot2) 
library(readr) 
library(dplyr)
library(stringr)
library(plotrix)
library(gridExtra)
library(RColorBrewer)
```  


```{r echo=FALSE}
sc_data<-read.csv("../input/Suicides in India 2001-2012.csv")
```
  
Structure of the dataset.   

```{r echo=FALSE}
str(sc_data)
summary(sc_data)
```  
Dataset contains data of suicide counts across the states of India, Age group, Gender,Year,type and type code.  
        
  
Who makes more suicide attempts Male or Female?  
```{r echo=FALSE}
colr<-c("dodgerblue4","mediumpurple3","goldenrod3","orangered4",
"lightsalmon4","mistyrose4","palevioletred3","slateblue4","slateblue","slategray4","tan")
gper<-sc_data %>% select(Gender,Total)%>% group_by(Gender)%>% summarise(total_all=sum(Total))%>%mutate(rs=sum(total_all), percent=round((total_all/rs)*100))
label <-  
  c( paste(gper$Gender[1],gper$percent[1],'%',sep=''),
     paste(gper$Gender[2],gper$percent[2],'%',sep=''))

ggplot(gper,aes(x=Gender,y=percent,fill=Gender))+geom_bar(stat="identity")+scale_fill_manual(values=c("dodgerblue4","goldenrod3"))+geom_text(aes(label=percent))
pie3D(gper$percent,labels=label,labelcex=1.1,explode=0.1,col=colr  	)

```
  
Which Age group makes more suicide attempts?  
```{r echo=FALSE}
sc_data %>% select(Gender,Age_group,Total)%>% filter(!Age_group=="0-100+")%>% group_by(Gender,Age_group)%>% summarise(atot=sum(Total))%>% ggplot(aes(x=Age_group,y=atot,fill=Gender))+geom_bar(stat="identity",position="dodge")+
scale_fill_manual(values=c("dodgerblue4","goldenrod3"))+labs(y="Count")



pdata<-sc_data %>% select(Age_group,Total)%>% filter(!Age_group=="0-100+")%>%group_by(Age_group)%>% summarise(atot=sum(Total))
pie3D(pdata$atot,labels=pdata$Age_group,explode=0.1,col=colr,
  	main="Suicide and Age Groups ")
```

```{r echo=FALSE}

sc_data %>% select(Age_group,Total)%>% filter(!Age_group=="0-100+")%>%group_by(Age_group)%>% summarise(atot=sum(Total))%>% ggplot(aes(x=Age_group,y=atot,fill=Age_group))+geom_bar(stat="identity")+
scale_fill_manual(values=colr)+labs(y="Count")
```


  
Education level and suicide  
```{r echo=FALSE} 
sc_type<-sc_data %>% filter(Type_code =="Education_Status")%>% select(Gender,Total,Type)%>% group_by(Gender,Type)%>% summarise(ttotal=sum(Total))
sc_type %>% ggplot(aes(x=str_sub(Type,1,15),y=ttotal,fill=Type))+geom_boxplot()+scale_fill_manual(values=colr)+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(x="Education Level",y="count")
```  

```{r echo=FALSE}
sc_type %>% ggplot(aes(x=str_sub(Type,1,15),y=ttotal,fill=Gender))+geom_bar(stat="identity",position="fill")+scale_fill_manual(values=colr)+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(x="Education Level",y="count")
```  
Clearly we can see education levels of Primary, Middle got more suicide counts.  
Academic performance and pressures have been found to be strongly related to teenage suicide problem and are one of the main causes leading to it.  
Also Uneducated poeple also attempts for it. People who are post graduate, diploma, graduate counts seems comparitively less.  
Education is the most powerful weapon which can change the world. School should do proper mentionring to the students.  
  
##Social Status of suicide Victims   
social status has got major influence in suicide attempts  
```{r echo=FALSE}
ss_type<-sc_data %>% filter(Type_code =="Social_Status")%>% select(Gender,Total,Type,Age_group)%>% group_by(Gender,Type,Age_group)%>% summarise(ttotal=sum(Total))
ss_type%>%ggplot(aes(x=Type,y=ttotal,fill=Type))+geom_boxplot()+scale_fill_manual(values=colr)+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(x="Social Status" ,y="count")
```  
  
Married people are the major victims of suicide comparing to other category.  

  
```{r echo=FALSE}
#ss_type %>% ggplot(aes(x=str_sub(Type,1,15),y=ttotal,fill=Gender))+geom_bar(stat="identity")+scale_fill_manual(values=colr)+coord_polar(theta = "y")+ geom_text(data = ss_type, hjust = 1, size = 3, aes(x = Type, y = 0, label = Type))+labs(x="",y="") 
```  
##Professional Profile  
Housewife are the major victims in suicide and next one was the people wo do farming and Agriculture.  
```{r echo=FALSE}
pp_type<-sc_data %>% filter(Type_code =="Professional_Profile")%>% select(Gender,Total,Type,Age_group)%>% group_by(Gender,Type,Age_group)%>% summarise(ttotal=sum(Total))
pp_type %>% ggplot(aes(x=str_sub(Type,1,20),y=ttotal,fill=Gender))+geom_bar(stat="identity",position="fill")+scale_fill_manual(values=colr)+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(x="Social Status",y="Percent")
```

```{r}

pp_type%>%ggplot(aes(x=Type,y=ttotal,fill=Type))+geom_bar(stat="identity")+scale_fill_manual(values=colr)+
theme(legend.position = "none",axis.text.x = element_text(angle=90))+labs(x="Professional Profile" ,y="count")
```
    
##Means Adopted  

```{r echo=FALSE}

ma_type<-sc_data %>% filter(Type_code =="Means_adopted") %>%group_by(Type,Gender,Age_group)%>%summarize(mtot=sum(Total))
ma_type%>%ggplot(aes(x=Type,y=mtot,fill=Gender))+geom_bar(stat="identity",position="dodge")+scale_fill_manual(values=colr)+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(x="Means Adopted",y="Count")
```  


```{r echo=FALSE}
ma_type%>%ggplot(aes(x=Type,y=mtot,fill=Age_group))+geom_bar(stat="identity",
                                                              position="dodge")+scale_fill_manual(values=colr)+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(x="Means Adopted",y="Count")

```  
    
Hanging is the most frequently used method of suicide.  

Materials for hanging were easily accessed and respondents considered it 'simple' to perform without the need for planning or technical knowledge.  
Hanging was thus seen as the 'quickest' and 'easiest' method with few barriers to completion and sometimes adopted despite not being a first choice.  

Prevention strategies should focus on countering perceptions of hanging as a clean, painless and rapid method that is easily implemented.  
However, care is needed in the delivery of such messages as some individuals could gain information that might facilitate fatal implementation.  
  
    
##Major Causes  
```{r echo=FALSE}

ca_type<-sc_data %>% filter(Type_code =="Causes")%>% group_by(Type,Gender,Age_group)%>%summarise(ctot=sum(Total))
ca_type%>%ggplot(aes(x=Type,y=ctot,color=Gender))+geom_point()+scale_color_manual(values=colr)+theme(legend.position = "bottom",axis.text.x = element_text(angle=65,vjust=0.5))+labs(x="Causes",y="Count")

```
   
At all age groups ** family Problems** seems to be the major cause for suicide, next stands prolonged illness/mental illness.  
At age group of 0-14, failure in examination was one of the main cause.  
```{r echo=FALSE}
options(repr.plot.width=6, repr.plot.height=4) 
p1<-ca_type%>% filter(Age_group =="0-14" & ctot!=0)%>% ggplot(aes(x=Type,y=ctot,fill=Type))+geom_bar(stat="identity")+theme(legend.position = "none",axis.text.x = element_text(angle=90,vjust=0.5))+labs(x="Major causes-Age 0-14",y="Count")
p2<-ca_type%>% filter(Age_group =="15-29" & ctot!=0)%>% ggplot(aes(x=Type,y=ctot,fill=Type))+geom_bar(stat="identity")+theme(legend.position = "none",axis.text.x = element_text(angle=90,vjust=0.5))+labs(x="Major causes-Age 15-29",y="Count")
p3<-ca_type%>% filter(Age_group =="30-44" & ctot!=0)%>% ggplot(aes(x=Type,y=ctot,fill=Type))+geom_bar(stat="identity")+theme(legend.position = "none",axis.text.x = element_text(angle=90,vjust=0.5))+labs(x="Major causes-Age 30-44",y="Count")
p4<-ca_type%>% filter(Age_group =="45-59" & ctot!=0)%>% ggplot(aes(x=Type,y=ctot,fill=Type))+geom_bar(stat="identity")+theme(legend.position = "none",axis.text.x = element_text(angle=90,vjust=0.5))+labs(x="Major causes-Age 45-59",y="Count")
p5<-ca_type%>% filter(Age_group =="60+" & ctot!=0)%>% ggplot(aes(x=Type,y=ctot,fill=Type))+geom_bar(stat="identity")+theme(legend.position = "none",axis.text.x = element_text(angle=90,vjust=0.5))+labs(x="Major causes-Age 60+",y="Count")
p1
p2
p3
p4
p5
```
  
Analysing the trend for the major causes for the suicides. Left the Reason unknown and Other causes, which are 2 mjor causes but it didn't give any specific meaning.  
  

```{r echo=FALSE}
sc_data %>% filter(Type_code=="Causes" & Type %in% c("Failure in Examination","Family Problems","Other Prolonged Illness","Unemployment","Dowry Dispute","Poverty","Insanity/Mental Illness"))%>%select(Year,Total,Type)%>% group_by(Year,Type)%>%summarise(ytot=sum(Total))%>%  ggplot(aes(x=factor(Year),y=ytot,color=Type,group=Type))+geom_line(size=1)+scale_color_manual(values=colr)+
theme(legend.position = "bottom",axis.text.x = element_text(angle=65,vjust=0.5))+labs(x="Year",y="Count")+geom_point(size=2)
```  
###Sucicides per Year
```{r exho=FALSE}
sc_data$Year<-as.factor(sc_data$Year)
sc_data%>%select(Year,Total,Age_group) %>% group_by(Year,Age_group) %>% summarise(total=sum(Total))%>%ggplot(aes(x=Year,y=total,group=Age_group,fill=Age_group))+
  geom_area()+scale_fill_manual(values=colr)
```  
  
    
Which state in India has got highest number of suicides.  
```{r echo=FALSE}
sc_data%>%select(State,Year,Total) %>% group_by(State,Year)%>% summarise(tot=sum(Total)) %>% ggplot(aes(x=State,y=tot,fill=Year))+geom_bar(stat="identity")+
theme(legend.position="bottom",axis.text.x=element_text(angle=90))

sc_data %>%filter(!State %in% c("Total (All India)","Total (States)","Total (Uts)"))%>%ggplot(aes(x=factor(Year),y=State,fill=Total))+geom_tile()+facet_wrap(~Gender)+scale_fill_gradientn(colors=brewer.pal(5,"Reds"))+theme(axis.text.x = element_text(angle=90))+labs(x="Year")
```
  
Finding the top 10 and bottom 10 states in india based on the number of suicides.  
```{r echo=FALSE}
options(repr.plot.width=6, repr.plot.height=4) 
sc1<-sc_data%>%filter(!State %in% c("Total (All India)","Total (States)","Total (Uts)"))%>% select(State,Year,Total) %>% group_by(State)%>% summarise(tot=sum(Total)) %>% arrange(desc(tot))%>%head(10)%>%
  ggplot(aes(x=factor(State,level=State),y=tot,color=State))+geom_point(size=4)+geom_segment(aes(xend=State,y=0,yend=tot),size=2)+theme(legend.position="none",axis.text.x=element_text(angle=90))+scale_color_manual(values=colr)+geom_text(aes(label=tot),vjust=0.3)+labs(x="State - More Suicides")
sc2<-sc_data%>%filter(!State %in% c("Total (All India)","Total (States)","Total (Uts)"))%>% select(State,Year,Total) %>% group_by(State)%>% summarise(tot=sum(Total)) %>% arrange(desc(tot))%>%tail(10)%>%
  ggplot(aes(x=factor(State,level=State),y=tot,color=State))+geom_point(size=4)+geom_segment(aes(xend=State,y=0,yend=tot),size=2)+theme(legend.position="none",axis.text.x=element_text(angle=90))+scale_color_manual(values=colr)+geom_text(aes(label=tot),vjust=0.3)+labs(x="State - Less Suicides")
grid.arrange(sc1,sc2,nrow=2)
```  
Maharastra is the State in Which has highest number of suicides, Next comes West Bengal,Tamilnadu,Andrapradesh and Karnataka.  
So we have to find what was the main causes in those states.  

Farming/Agriculture is the major occupation across all states of India, due of shortage of rainfall. Farming was impacted more which causes more suicide attempts.  
Finding the states which has more counts.  
```{r echo=FALSE}
st_mh<-sc_data %>% filter( State=="Maharashtra" | (Type_code=="Professional_Profile" & Type=="Farming/Agriculture Activity"))%>% select(Age_group,State,Year,Total)%>% group_by(State,Year)%>% summarise(stot=sum(Total))%>% arrange(desc(stot))
st_mh%>%ggplot(aes(x=State,y=stot,group=Year,fill=Year))+geom_area()+theme(legend.position="bottom",axis.text.x=element_text(angle=90))+scale_color_manual(values=colr)
```
  
  
Sexual abuse is one of leading factor for suicidal thoughts and attempts all over the World.  
```{r echo=FALSE}
sc_data %>% filter (Type=="Physical Abuse (Rape/Incest Etc.)")%>% select(Year,State,Age_group,Total)%>% group_by(Year,Age_group)%>%
summarise(ptot=sum(Total))%>% ggplot(aes(x=factor(Year),y=ptot,group=Age_group,fill=Age_group))+geom_area()+labs(x="Year",y="Count",fill="Age Group")+scale_fill_manual(values=colr)

```  
Impact was more in the Age groups of 0-14 and 15-29.